Climate-driven Phytoplankton community variability in the Cariaco Basin, Venezuela

### 2025-02-23 agujournaltemplate.tex
Focus Phase Day 7 Commit - restructure to final draft of Intro

    Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}. These communities consist of a diverse set of species that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.

    %DIVERSITY PARAGRAPH HERE? yes...

    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}. 
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}. 
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test-bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Instead of going paper by paper below, I should go topic by topic (i.e. all points made on "DIVERSITY" should be in a paragraph, and all on changes in FGs or Size Classes in another etc...)

@@

    The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts and high performance liquid chromatography (HPLC) pigment analysis. Of these, all cover the entire time series except the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.

    - Taylor 2012 ()
    \citeA{taylor_ecosystem_2012} analyzed data up until 2010 and found significant decadal trends of sea surface temperature (SST) increase, upwelling intensity decrease and diminished phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified via microscopy showed a shift towards smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass. The authors explained the change in physical regime in the Cariaco basin with a north-easterly shift of the Inter-Tropical Convergence Zone (ITCZ) and a resulting weakening trend in Trade Winds.

@@

    - Finally what it says in Review about Phytoplankton community
    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!

@@

@@

- We are analyzing long term time series of data, where probably different analysts have counted the phytoplankton samples...
%It is well known that the personal component has an influence on how well species are identified. Even the same person counting the samples can evolve over time, gaining more experience, which can also influence the results (Löder et al. 2012; Nohe et al. 2018). A recommendation for further improvement of the assessment is a case wise cross-validation of some samples by the monitoring agencies involved. Another important issue to be considered are the differences in sampling frequency between countries and also between stations (see ESM Fig. S1). In highly dynamic and turbid waters such as the Wadden Sea, low frequency and infrequent sampling is not sufficient to properly capture temporal signals (Fettweis et al. 2023). % from Di Cavalho et al. 2023

>>>>

    % Phytoplankton is important
    Phytoplankton communities provide the foundation of marine food webs and play a crucial role in global biogeochemical cycles \cite{falkowski_biogeochemical_1998}.
    Phytoplankton consists of a diverse set of photoautotrophic microorganisms that exhibit highly variable ecophysiological properties \cite{appeltans_magnitude_2012}. The short generation time and high diversity allow phytoplankton communities to adapt quickly to changes in the environment, making them a highly suitable indicator of changes in an ecosystem \cite{alvarez-cobelas_what_1998, di_cavalho_temporal_2023}.

    - Phytoplankton communities can be described from taxonomic/functional group aspects, but also from biodiversity aspects (or trait-based, but this I am not doing here)

    % Phytoplankton is under threat
    On a global scale, human activities are driving a decline in biodiversity \cite{tittensor_mid-term_2014}. Climate change is predicted to have a lasting impact on phytoplankton communities, shifting the ranges and contributions of functional groups \cite{henson_future_2021}.

    %%% Phytoplankton functional group shifts through climate change
    %%% Phytoplankton diversity shifts through climate change

    The loss of biodiversity is predicted to have a lasting impact on the functioning of an ecosystem, since links have been found between the richness of a phytoplankton community and the biomass and resource use efficiency \cite{ptacnik_diversity_2008, lehtinen_phytoplankton_2017}.

    - mention BDEF in Tropics paper, or some other paper, to highlight that the tropics are notoriously understudied in relation to phytoplankton diversity.

    %%% This paper is clearly not BDEF, but environmental impacts on diversity and community!section

    % We need time series in order to understand effects and changes (threats) 
    Time series of biological observations offer valuable insight into how marine ecosystems react to temporary shifts, cyclic processes, and long-term climate changes, especially when data collection spans several decades \cite{carstensen_need_2014, henson_observing_2016}.

    % CARIACO is one such time series
    One such project is the CARIACO time series, which has seen consistent sampling of hydrographic and biogeochemical parameters in the Cariaco basin between 1995 and 2017 \cite{muller-karger_scientific_2019}.

    % Describe CARIACO location
    The Cariaco Basin, located off the coast of Venezuela in the southern Caribbean Sea, is a highly productive ecosystem within the largest truly marine anoxic basin in the world \cite{edgcomb_accessing_2011}. A markedly seasonal upwelling cycle is driven by easterly trade winds, with the highest levels of primary production observed from December through May, and a smaller secondary peak in July and August \cite{mullerkarger_annual_2001, astor_seasonal_2003}. The basin is a continental shelf depression consisting of two sub-basins of \qty{\sim 1400}{m} depths, which are effectively separated from the Caribbean Sea by a sill with a mean depth of about \qty{100}{m}. The water below the sill is poorly ventilated, which, combined with the high surface productivity and microbial respiration of sinking organic matter, results in permanent anoxic conditions at depths below \qty{\sim 275}{m} \cite{thunell_organic_2000}. Due to these unique properties, the Cariaco basin has been used as a natural laboratory since the 1950s, providing a rich sedimentary record for the study of the past climate \cite{hughen1996nature} and a test bed for studying the coupling between pelagic production and carbon export to the deep ocean \cite{montes_vertical_2012}.

    % Describe CARIACO changes in ecosystem

    %% Changes in Functional Groups

    %% Changes in Diversity

    % Now we come into the picture: Which gap are we adressing?

    %% in depth look at biodiversity aspect, in particular microscopy data

    %% try to identify drivers of shift in community with Gradient Forest

    %%% Instead of going paper by paper below, I should go topic by topic (i.e. all points made on "DIVERSITY" should be in a paragraph, and all on changes in FGs or Size Classes in another etc...)

@@

    The time series includes multiple measurements of the phytoplankton community, which are the bulk fluorometric chlorophyll, microscopy cell counts, and high-performance liquid chromatography (HPLC) pigment analysis. Of these, all cover the entire time series except for the HPLC pigment data, which was performed by three different labs and with a gap between 2000 and 2006 \cite{muller-karger_scientific_2019}.

    The most often measured indicator of bulk phytoplankton biomass is fluorometric Chlorophyll \textit{a} concentration, which gives a general measure of biomass standing stock and distribution in the water column. A more advanced methodology, high-performance liquid chromatography (HPLC), allows individual pigment types to be quantified. These pigment distributions can be used to estimate the groups of phytoplankton that contribute to biomass. The most specific methodology to quantify phytoplankton communities is the counting and identification of individual phytoplankton cells, either manually under a microscope or automated in flow-through systems. The major drawbacks are the size limit of resolving smaller cells and, if done manually, the complexity of identifying phytoplankton species.

    % PAPER TIMELINE:
    - Taylor 2012 ()
    \citeA{taylor_ecosystem_2012} analyzed data up to 2010 and found significant decadal trends of increase in sea surface temperature (SST), coinciding with a decrease in upwelling intensity and phytoplankton bloom intensities, which the authors related to the impact of global warming. The phytoplankton community identified via microscopy showed a shift towards smaller taxa after 2004, which coincided with a collapse in commercial sardine landings and an increase in mesozooplankton biomass. The authors explained the change in physical regime in the Cariaco basin with a north-easterly shift of the Inter-Tropical Convergence Zone (ITCZ) and a resulting weakening trend in Trade Winds.

@@

    - Finally, says in Review about Phytoplankton community
    - Understanding the dynamics of the Cariaco Basin is crucial to understanding the broader ecological changes in tropical marine ecosystems. + Carbon fluxes in general!

@@

    % INTRO TEXT DUMP:

    Biodiversity is a complex concept composed of genetic, taxonomic, and ecological aspects that can additionally be quantified in temporal and spatial scales. In its basic sense, a measure of diversity should quantify two components: the number of items and their relative frequencies \cite{borics_freshwater_2021}. 
    Richness, the number of species or genera present at a particular time in a sample, has been established as the most basic descriptor of community diversity, but it is not the best indicator of ecosystem change \cite{hillebrand_biodiversity_2018}.
    - Other metrics, evenness, and dominance

@@

- We are analyzing long term time series of data, where probably different analysts have counted the phytoplankton samples...
%It is well known that the personal component has an influence on how well species are identified. Even the same person counting the samples can evolve over time, gaining more experience, which can also influence the results (Löder et al. 2012; Nohe et al. 2018). A recommendation for further improvement of the assessment is a case wise cross-validation of some samples by the monitoring agencies involved. Another important issue to be considered are the differences in sampling frequency between countries and also between stations (see ESM Fig. S1). In highly dynamic and turbid waters such as the Wadden Sea, low frequency and infrequent sampling is not sufficient to properly capture temporal signals (Fettweis et al. 2023). % from Di Cavalho et al. 2023
- Discuss limitation of taxonomic diversity, i.e. not direct translation to functional diversity, but here no Data foundation to look at functional diversity.. (inspired by Borics et al. 2021)
